A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-22 paper.tex
Apply Gregors comments

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $D_3$ and $D_4$.
Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the product of
recombining $D_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
This recombination at means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
these intervals are shown attached to the corresponding graph edges.

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
Thus, we have the means of encoding the effects of
multiple \emph{simultaneous} events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

Genome ARGs, on the hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
% Wilder: finds the cell-to-cell confusing. Maybe drop?
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

Ancestral material is a critical concept in a retrospective ARG,
and refers to the genomic intervals directly ancestral to the samples
scattered across lineages (the edges of the graph)
at some time in the past. At recombination
events, ancestral material is split between the ancestral lineages.
As genome segments coalesce in a common
ancestor the total amount of ancestral material is reduced, until
eventually, as we traverse rootwards through the graph, all
samples reach a most recent common ancestor at every location
along the genome. The process of transmitting ancestral material
along the edges of an ARG is directly analogous to Hudson's
simulation algorithm for the coalescent with
recombination~\citep{hudson1983testing,kelleher2016efficient}.

@@

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a therefore a
unary node in these three trees.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]

@@

\cite{shipilina2023origin} discuss the same ideas, and note that
``full ARG ... contains more information than the series of tree
sequences along the genome''. They view the ``full ARG'' as an
eARG, and the ``series of tree sequences'' is their
% jk: probably should delete this. It *is* muddled though
(somewhat muddled)
term for the sequence of local trees along the genome. We can see
that their statement is true, \emph{if} the local trees contain
no unary nodes. However, as we have shown, if the local trees
contain labelled unary nodes, then  there is precisely as
much information in them as an ARG. Whether such unary nodes
are \emph{observable} is an entirely separate question,
of course.

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
meaning that sample size is substantially larger than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination, but which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.

>>>>

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, diploid individual $D_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $D_3$ and $D_4$.
Genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the recombinant product of
$D_2$'s two genomes \noderef{e} and \noderef{f} crossing over at position 2.
This recombination means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and $[2, 10)$ from \noderef{f}.
These intervals are shown attached to the corresponding graph edges.

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
[NOTE: we could move this to the later section currently
called "the equivalence of ARGS and the local trees",and
refocus a bit to "ARGs and local trees".]
The relationship between an ARG and the corresponding local trees
is subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means of encoding the effects of
multiple \emph{simultaneous} events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
Genetic inheritance must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

Genome ARGs, on the hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
% Wilder: finds the cell-to-cell confusing. Maybe drop?
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

Ancestral material is a critical concept in a retrospective ARG,
and refers to the genomic intervals ancestral to the samples
scattered across lineages (the edges of the graph)
at some time in the past. At recombination
events, ancestral material is split between the ancestral lineages.
As genome segments coalesce in a common
ancestor the total amount of ancestral material is reduced, until
eventually, as we traverse rootwards through the graph, all
samples reach a most recent common ancestor at every location
along the genome. The process of transmitting ancestral material
along the edges of an ARG is directly analogous to Hudson's
simulation algorithm for the coalescent with
recombination~\citep{hudson1983testing,kelleher2016efficient}.

@@

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of a marginal distributions indicating
an integrated tree over many positions.

@@

\cite{shipilina2023origin} discuss the same ideas, and note that
``full ARG ... contains more information than the series of tree
sequences along the genome''. They view the ``full ARG'' as an
eARG, and the ``series of tree sequences'' is their
% jk: probably should delete this. It *is* muddled though
(somewhat muddled)
term for the sequence of local trees along the genome.
Their statement is true, \emph{if} the local trees are
unlabelled and contain
no unary nodes. However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG. Whether such unary nodes
are \emph{observable} is an entirely separate question,
of course.

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
meaning that sample size is substantially larger than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Finally, the COVID-19 pandemic has provided an example of a dataset that
is clearly a very poor fit for the coalescent with recombination,
but for which,
nevertheless, we would like to estimate ARGs.
\cite{zhan2023towards} inferred an ARG for 1.27 million SARS-CoV-2 whole
genomes, and showed how it accurately captures several known properties of
the pathogen's evolution, and can efficiently and concisely
encode this vast dataset.
